\bfseries\LARGE\singlespacing{Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols}

### 2023-12-18 doc/comparison-mcid12/manuscript.tex
Updates from Overleaf

\usepackage{import}
\usepackage{../paper/main-lancet}

@@

\title{\bfseries\LARGE\singlespacing{Brief report:
Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols}}

@@

\begin{information}\normalfont
\textbf{Word count:} Abstract 150 (max 150), Manuscript 1,625 (max 1,500) \\
\textbf{Display items:} Figure 1 \\
\textbf{References:} 17 (max 12) \\
%\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
% \par
\end{information}

\begin{abstract}\normalfont
\noindent We compared airborne and saliva samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting. While during the pandemic we detected almost exclusively SARS-CoV-2 in saliva (19~samples of SARS-CoV-2 vs 2~samples of non-SARS-CoV-2), after the pandemic the most common respiratory viruses were adenovirus, influenza, and rhinoviruses (3~samples of SARS-CoV-2 vs 47~samples of non-SARS-CoV-2). Despite the same study setting and methods used, airborne detection of SARS-CoV-2 was relatively more frequent than of non-SARS-CoV-2 viruses (10~samples of SARS-CoV-2 in 2021/2022 vs 2~samples of non-SARS-CoV-2 in 2022/2023). SARS-CoV-2 was also detected more frequently on the HEPA-filters of installed portable air cleaners (6~samples of SARS-CoV-2 vs 3~samples of non-SARS-CoV-2). Our comparison suggests that SARS-CoV-2 is easier to detect in the air than other endemic respiratory viruses, possibly facilitating long-range transmission. \medskip

@@

Data was collected in Swiss secondary schools (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from end of January to beginning of March. Two classes of two different secondary schools (school 1 and 2) participated in 2022 and two classrooms of one secondary school (\ie only school 2) participated in 2023. In 2022, a general mask mandate was in place for the first 3~study weeks, followed by 2~weeks without intervention, and another 2~weeks during which air cleaners were installed in the classrooms. In 2023, air cleaners were installed for half of the study period in the classrooms following a cross-over study design.

All classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system, but were ventilated using passive window ventilation. An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, temperature, and humidity. In 2022, passive window ventilation occurred per recommendations of the national public health authorities. In school~1, ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top. In 2023, passive window ventilation occurred solely at the discretion of the teachers.

Receptive testing for a panel of respiratory infections was done weekly in 2022 and bi-weekly in 2023. Airborne respiratory viruses were collected in each classroom with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France) and a BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft. Collins, CO, USA)\cite{Lednicky2016AST}. The removable parts were regularly autoclaved. Finally, swabs from the air cleaners' HEPA filters were collected at the end of the study in 2022 and after each intervention phase in 2023. The HEPA filters were removed and divided into 20~fields. One sterile Phosphate-Buffered Saline-moistened swab per field was then taken for a total of 20~swabs per filter. Saliva and airborne samples were transported to the Institute of Infectious Diseases and stored at $-$80°C until further processing\cite{Huber2021}.

Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL. Saliva samples were analyzed directly without prior filtration. The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2, influenza A/B, respiratory syncytial, metapneumovirus, adenovirus, rhinovirus, and parainfluenza.

Descriptive statistics were used to present differences in the type, number, and proportion of respiratory viruses detected in saliva and airborne samples between 2022 and 2023.

@@

In 2022, 44~students participated in weekly saliva testing. There were 21~positive saliva samples over the study, namely 19~SARS-CoV-2, 1~influenza~A, and 1~adenovirus. There were 10~positive bioaerosol samples, namely 9~SARS-CoV-2 and 1~adenovirus. There were 8~positive samples on the HEPA-filters of the air cleaners, namely 6~SARS-CoV-2, 1~influenza~A and 1~adenovirus. In 2023, 38~students from two classes of secondary school~2 participated in bi-weekly saliva testing. There were 50~positive saliva samples over the study, namely 15~influenza~B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, 1~parainfluenza virus. There were 2~positive bioaerosol samples, namely 1~rhinovirus and 1~adenovirus. There were 4~positive samples on the HEPA-filters of the air cleaners, namely 1~influenza~B, 1~rhinovirus, 1~adenovirus, and 1~SARS-CoV-2.

In comparison, there was a clear shift in the distribution of respiratory viruses from predominantly SARS-CoV-2 in 2022 to predominantly non-SARS-CoV-2 viruses in 2023 (\Cref{fig:comparison}a). Molecular detection also varied between the two years, especially for airborne samples (\Cref{fig:comparison}b). For saliva, 21 out of 262~samples were positive in 2022 compared to 50 out of 448~samples in 2023. For bioaerosols, 10 out of 130~samples were positive in winter 2022 compared to 2 out of 105~samples in 2023 . For the HEPA-filters, 8 out of 40~swabs were positive in 2022 compared to 4 out of 160~swabs in 2023.

\begin{figure}
    \centering
    \includegraphics{results/comparison.pdf}
    \caption{Comparison of molecular detection of respiratory viruses between winter 2021/2022 (2022) and winter 2022/2023 (2023). \textbf{(a)}~Distribution of respiratory viruses found in saliva. IF: influenza~A/B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus. \textbf{(b)}~Number of positive samples in saliva, bioaerosols, and on the HEPA-filters of air cleaners.}
    \label{fig:comparison}
\end{figure}

Daily mean CO$_2$ level was 1,064\,ppm (standard deviation [SD] 232\,pppm) in 2022 compared to 1,702\,ppm (SD 370\,pppm) in 2023, suggesting that the classrooms were less well ventilated after the COVID-19 pandemic. Daily mean relative humidity was 38\% (SD 6\%) in 2022 compared to 38\% (SD 5\%) in 2023. Daily mean temperature was 19°C (SD 2°C) in 2022 compared to 22°C (SD 1°C) in 2023.

@@

We compared molecular data collected from two studies in Swiss secondary schools in winter 2021/2022 (2022) and 2022/2023 (2023). During the COVID-19 pandemic in 2022, we detected mainly SARS-CoV-2 in saliva and also in bioaerosols and on the HEPA-filters of air cleaners. After the pandemic in 2023, we detected a clear shift in the distribution of respiratory viruses towards mainly non-SARS-CoV-2 respiratory viruses such as influenza, adenovirus, and rhinovirus. These viruses were rarely detected in bioaerosols and on the HEPA-filters of air cleaners.

The differences between the two study periods (pandemic versus non-pandemic) suggest that SARS-CoV-2 is easier to detect in the air. One possible explanation is that SARS-CoV-2 can remain suspended and active in the air for extended periods of time, facilitating long-range transmission. In contrast, non-SARS-CoV-2 viruses, such as influenza and adenovirus, were rarely detected in the air, suggesting they settle or become inactive more quickly, and require prolonged close contact for transmission. However, these findings should be interpreted with caution as they originate from an analysis of secondary data collected in a real-world setting, not a rigorous experiment. Furthermore, although there may be variations in the airborne survival of respiratory viruses, it is important to note that prolonged close contact continues to play a significant role in transmission. This holds true even for viruses capable of long-range transmission, as exemplified by SARS-CoV-2\cite{Leung2020NatMed,Lind2023NatCommun}.

Technical reasons for the differences in airborne detection are unlikely because the bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed. Instead, airborne detection could have been influenced by environmental factors. Although temperature and relative humidity were similar in both study periods, the same levels may have different effects on the airborne survival of respiratory viruses\cite{Tellier2009JTRSI,Davis1971AM}. A change in ventilation conditions could also have had a profound effect, but CO$_2$ levels were even higher in winter 2023, suggesting that classrooms were not as well ventilated as during the COVID-19 pandemic. We speculate that the difference in airborne molecular detection is due to the different characteristics of the respiratory viruses. Differences in the physicochemical properties of virus-laden aerosols (\eg particle size and viral load), which influence the generation and survival of airborne viral RNA\cite{Wang2021}, may facilitate the detection of SARS-CoV-2 in the air. This is because viral loads often do not exceed the detection limits of existing devices for bioaerosol sampling, which remains a challenge\cite{Belser2023PLOSPath}. For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}. However, several studies , including ours, have shown that sporadically non-SARS-CoV-2 viruses such as influenza, rhinovirus, and adenovirus can also be detected in the air\cite{Bischoff2013JID,Pan2017mSphere,Myatt2004AJRCCM}. Finally, the ability to detect pathogens in bioaerosols may be related to the patient characteristics of the study participants. Several studies have suggested an association between the generation of pathogenic aerosols and infectiousness\cite{Leung2020NatMed,Bischoff2013JID}. Therefore, studies that include more infectious patients (intentionally or unintentionally) are likely to have a higher chance of detecting respiratory viruses in the air.

In conclusion, our comparison of two years of respiratory viral transmission in a school setting suggests that SARS-CoV-2 is easier to detect in indoor air as compared to other endemic viruses such as influenza, adenovirus and rhinoviruses. Future studies should test this hypothesis experimentally and investigate the relative importance of different modes of transmission.

% We detected fewer respiratory viruses in bioaerosol samples compared to our previous study\cite{Banholzer2023PLoSMed} (1~sample of adenovirus and 1~sample of rhinovirus in this study vs 10~samples of SARS-CoV-2 in the previous study), especially in relation to the number of positive saliva samples (50~samples in this study vs 21~in the previous study). Similarly, the number of viruses detected on the filters of the air cleaners was also lower in this study (4~positive samples in 160~swabs vs 8~positive samples in 80~swabs in the previous study). Importantly,  in the current study compared to the previous study in 2022 during the SARS-CoV-2 Omicron variant. Whereas in 2022 we detected almost exclusively SARS-CoV-2 in human saliva, the most common respiratory viruses in 2023 were adenovirus, influenza~B, and rhinoviruses (only three SARS-CoV-2-positive samples). The shift in the pattern of respiratory viruses has been observed in other studies\cite{Nygaard2023Lancet,Sauteur2022EuroSurv}.

>>>>

\usepackage{import}
\usepackage{main}

@@

\title{\bfseries\LARGE\singlespacing{Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols}}

@@

\begin{information}\normalfont
\textbf{Word count:} Abstract 219 (max 250), Manuscript 1,380 (max 1,200) \\
\textbf{Display items:} Figure 1 (max 2) \\
\textbf{References:} 15 (max 15) \\
%\textbf{Supporting information:} STROBE checklist, Supplementary Data with texts, tables, and figures
% \par
\end{information}

\begin{abstract}\normalfont
\noindent \textbf{Objectives:} To compare molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and bioaerosols.

\noindent \textbf{Methods:} We analyze saliva and bioaerosol samples from two studies conducted in winter 2021/2022 (during the SARS-CoV-2 omicron wave) and winter 2022/2023 (after the COVID-19 pandemic) in a Swiss school setting. We estimate the probability of weekly airborne detection of SARS-CoV-2 vs non-SARS-CoV-2 viruses using a Bayesian logistic regression model.

\noindent \textbf{Results:} While during the pandemic SARS-CoV-2 was almost exclusively detected in saliva (19~samples of SARS-CoV-2 vs 2~samples of non-SARS-CoV-2), the most common respiratory viruses after the pandemic were adenovirus, influenza, and rhinoviruses (3~samples of SARS-CoV-2 vs 47~samples of non-SARS-CoV-2). Despite the same study setting and methods used, airborne detection of SARS-CoV-2 was relatively more frequent than of non-SARS-CoV-2 viruses (10~samples of SARS-CoV-2 in 2021/2022 vs 2~samples of non-SARS-CoV-2 in 2022/2023). Based on our model, the posterior probability of weekly airborne detection was 31\% (95\%-credible interval [CrI] 13\%$-$54\%) for SARS-CoV-2 vs 11\% (95\%-CrI 4\%-23\%) for non-SARS-CoV-2.

\noindent \textbf{Discussion:} There was a clear shift in the distribution of respiratory viruses from predominantely SARS-CoV-2 in 2022 to other respiratory viruses in 2023. Our comparison suggests that SARS-CoV-2 is easier to detect in the air than other endemic respiratory viruses, possibly facilitating long-range transmission. Future experimental studies are needed to confirm this finding. \medskip

@@

Data was collected in two Swiss secondary schools (age of students 14-17~years) in the canton of Solothurn, Switzerland, over a seven-week study period from end of January to beginning of March. Three classes participated in 2022 and two classes in 2023. Two classes shared the same classroom in 2022 due to half-class teaching, other classes had a separate classroom. All classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system, but were ventilated using passive window ventilation. An air quality device (AQ Guard, Palas GmbH, Karlsruhe, Germany) continuously measured indoor CO$_2$ levels, temperature, and humidity. A detailed comparison of the study settings can be found in \supp~Table~\zref{tab:comp_study}.

Repetitive testing for a panel of respiratory infections was done weekly in 2022 and bi-weekly in 2023. Airborne respiratory viruses were collected in each classroom with a cyclonic bioaerosol sampling device (Coriolis Micro Air, Bertin Instruments Montigny-le-Bretonneux, France) and a BioSpot-VIVAS condensation particle growth collection device (Aerosol Devices Inc., Ft. Collins, CO, USA)\cite{Lednicky2016AST}. The removable parts were regularly autoclaved. Finally, swabs from the air cleaners' HEPA filters were collected. The HEPA filters were removed and divided into 20~fields. One sterile Phosphate-Buffered Saline-moistened swab per field was then taken for a total of 20~swabs per filter. Saliva and airborne samples were transported to the Institute of Infectious Diseases and stored at $-$80°C until further processing\cite{Huber2021}. Prior to the real-time (RT)-PCR analysis, daily bioaerosol samples were combined for each sampling device and filtered using Amicon Ultra-15 Centrifugal Filters with Ultracel 10,000 Dalton molecular weight cutoffs filters (UFC9010; MilliporeSigma, Burlington, USA) to a volume of 1\,mL. Saliva samples were analyzed directly without prior filtration. The Allplex RV Master Assay (Seegene, Seoul, South Korea) detects a panel of 19 major respiratory viruses and viral subtypes, including SARS-CoV-2, influenza A/B, respiratory syncytial, metapneumovirus, adenovirus, rhinovirus, and parainfluenza.

Descriptive statistics were used to present differences in the type and number of respiratory viruses detected in saliva and airborne samples between 2022 and 2023. A Bayesian logistic regression model was used to estimate the probability of detecting any SARS-CoV-2 vs non-SARS-CoV-2 in the air during a study week, adjusting for differences in the study settings and the interventions that were in place during the studies (see \supp~Text~\zref{sec:model} for a detailed model description).

@@

In 2022, 58~students participated in weekly saliva testing. There were 21~positive saliva samples over the study, namely 19~SARS-CoV-2, 1~influenza~A, and 1~adenovirus (\Cref{fig:comparison}a, left). There were 10~positive bioaerosol samples, namely 9~SARS-CoV-2 and 1~adenovirus. There were 8~positive samples on the HEPA-filters of the air cleaners, namely 6~SARS-CoV-2, 1~influenza~A and 1~adenovirus. In 2023, 38~students participated in bi-weekly saliva testing. There were 50~positive saliva samples over the study, namely 15~influenza~B, 15~rhinovirus, 14~adenovirus, 3~SARS-CoV-2, 2~metapneumovirus, 1~parainfluenza virus (\Cref{fig:comparison}a, right). There were 2~positive bioaerosol samples, namely 1~rhinovirus and 1~adenovirus. There were 4~positive samples on the HEPA-filters of the air cleaners, namely 1~influenza~B, 1~rhinovirus, 1~adenovirus, and 1~SARS-CoV-2.

We found 6 positive air-saliva samples in the same respective week (4~SARS-CoV-2 and 2~non-SARS-CoV-2 (\Cref{fig:comparison}b), suggesting they were paired samples. Based on our Bayesian logistic regression model, SARS-CoV-2 was more likely detected in bioaerosols than non-SARS-CoV-2 (adjusted odds ratio 3.6, 95\%-CrI 1.5$-$8.9). The probability of molecular airborne detection was 31\% (95\%-credible interval [CrI] 13\%$-$54\%) for SARS-CoV-2 vs 11\% (95\%-CrI 4\%-23\%) for non-SARS-CoV-2 (\Cref{fig:comparison}c). Our estimates are adjusted for differences in the daily maximum CO$_2$ level, which was 1,134\,ppm (standard deviation [SD] 277\,pppm) in 2022 compared to 2,224\,ppm (SD 321\,pppm) in 2023. Humidity and temperature were similar, with daily mean relative humidity of 38\% (SD 6\%) in 2022 vs 38\% (SD 5\%) in 2023 and daily mean temperature of 19°C (SD 2°C) in 2022 vs 22°C (SD 1°C) in 2023.

\begin{figure}
    \centering
    \includegraphics{results/comparison.pdf}
    \caption{Comparison of molecular detection of respiratory viruses between winter 2021/2022 (2022) and winter 2022/2023 (2023). \textbf{(a)}~Distribution of respiratory viruses found in saliva. IF: influenza~A/B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus. \textbf{(b)}~Positive samples in saliva and bioaerosols per study week. \textbf{(c)}~Probability of detecting any SARS-CoV-2 and non-SARS-CoV-2 in bioaerosols during a study week (posterior mean as dots, interquartile range as box, 95\%-CrI as error bars).}
    \label{fig:comparison}
\end{figure}

@@

% In comparison,

We compared molecular data collected from two studies in Swiss secondary schools in winter 2021/2022 (2022) and 2022/2023 (2023). We detected predominantly SARS-CoV-2 in 2022 and predominantly non-SARS-CoV-2 viruses, such as influenza and adenovirus, in 2023. Molecular airborne detection of SARS-CoV-2 was more likely than of non-SARS-CoV-2.

The differences between the two study periods (pandemic versus non-pandemic) suggest that SARS-CoV-2 is easier to detect in the air. One possible explanation is that SARS-CoV-2 can remain suspended and active in the air for extended periods of time, facilitating long-range transmission. In contrast, non-SARS-CoV-2 viruses were rarely detected in the air, suggesting they settle or become inactive more quickly. Therefore, prolonged close contact may be relatively more important for the transmission of respiratory viruses other than SARS-CoV-2, although previous studies suggest it also facilitates the transmission of SARS-CoV-2\cite{Leung2020NatMed,Lind2023NatCommun}.

Technical reasons for the differences in airborne detection are unlikely because the bioaerosol sampling devices and laboratory procedures were the same in both studies, and no technical problems were observed. Temperature and relative humidity were also similar in both study periods, but the same levels may have virus-specific effects on airborne survival\cite{Tellier2009JTRSI,Davis1971AM}. Ventilation changed profoundly, but CO$_2$ levels were even higher in 2023, possibly enhancing airborne survival. Additional differences in the study setting (\supp~Table~\zref
{tab:comp_study}) that could have influenced airborne detection were adjusted for in the Bayesian model (\supp~Text~\zref{sec:model}). Therefore, we speculate that the difference in airborne detection is due to the different characteristics of the respiratory viruses which can influence the generation and survival of airborne viral RNA\cite{Wang2021}. Low viral loads do often not exceed the detection limits of existing devices for bioaerosol sampling. For example, one study failed to detect any human rhinovirus in the exhaled breath of 16~HRV-infected subjects\cite{Fabian2011JAMPDD}. Finally, the ability to detect pathogens in bioaerosols may be related to the variation in infectiousness between study participants, which is associated with the generation of pathogenic aerosols\cite{Leung2020NatMed,Bischoff2013JID}.

In conclusion, we observed a clear shift in the distribution of respiratory viruses from predominantly SARS-CoV-2 in 2022 to predominantly non-SARS-CoV-2 viruses in 2023. Molecular airborne detection of SARS-CoV-2 was more likely than of other endemic respiratory viruses, but future experimental studies are needed to confirm this finding.
